---
title: "activity tracker calculation"
author: "Daniel Spakowicz"
date: "5/28/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(MASS)
library(magrittr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(CausalImpact)
```

# ROAR-LCT 

20-30 participants ≥ 60 years of age with advanced lung cancer who plan to receive treatment from the OSUCCC Thoracic Oncology Clinic will be enrolled.

Activity tracker data will be used to monitor activity and resting heart rate. 

Below is an example dataset that includes two covariates of varying periodicities and a "state change", i.e. an effect of an intervention.

```{r example dataset}
set.seed(123456)

# length of the series
len <- 500
treatment <- ceiling(2*len/3) #beginning of the treatment
postperiod <- len - treatment

# parameters of the state
statedim <- 3
T <- diag(statedim) #parameter matrix of state
Q <- 0.01*diag(statedim) #variance of state
x1 <- sin(2*pi*seq_len(len)/90)
x2 <- sin(2*pi*seq_len(len)/360)

# simulation of the state
z0 <- c(5,0,0) #initial conditions: mu_0 = 5, beta_10 = 0, beta_20 = 0
z <- matrix(ncol = statedim, nrow = len)

# Start with a draw based on the initial conditions z0
z[1,] <- z0%*%T + mvrnorm(1,rep(0,statedim), Q)

# Update using the conditions from the last draw (previous column)
for (i in 2:len) {
  z[i,] <- z[(i-1),] %*% T + mvrnorm(1, rep(0, statedim), Q)
}

# observations
sigmasq <- 0.01 
Z <- matrix(ncol = statedim, nrow = len) 
Z[,1] <- rep(1, len)
Z[,2] <- x1
Z[,3] <- x2

e <- 0.25 #effect size
Ze <- Z
Ze[treatment:len, 1] <- Ze[treatment:len, 1] * (1 + e)

# Create a y vector without a state change
y <- vector(length = len)
for (i in 1:len) {
  y[i] <- Z[i,] %*% z[i,] + sqrt(sigmasq)*rnorm(1)
}

# Create y vector with a state change (effect of intervention)
ye <- vector(length = len)
for (i in 1:len){
  ye[i] <- Ze[i,] %*% z[i,] + sqrt(sigmasq)*rnorm(1)
}

# Create data frame containing all variables to be plotted
z.plot <- 
  Ze %>%
  data.frame %>%
  dplyr::rename(state = X1, x1 = X2, x2 = X3) %>%
  mutate("y (no state change)" = y) %>%
  mutate("y (with state change)" = ye) %>%
  mutate(index = seq(1, nrow(.)))

z.plot %>%
  gather(key, value, -index) %>%
  ggplot(aes(index, value)) +
  geom_line(aes(group = key, color = key)) +
  geom_vline(xintercept = treatment, lty = "dotted") +
  theme_bw() +
  labs(x = "Time", color = "") +
  ggsave("../figures/example_state_change.png", height = 4, width = 7.5)
```

## Power calculation

```{r}
ci.pwr <- 
  z.plot %>%
  dplyr::select("y (with state change)", x1, x2, index) %>%
  dplyr::rename("y" = "y (with state change)") %>%
  zoo

impact <- 
  CausalImpact::CausalImpact(ci.pwr, 
                           pre.period = c(1, treatment), 
                           post.period = c(treatment + 1, 
                                           treatment + postperiod)
)

plot(impact) +
  ggsave("../figures/example_impact.png", height = 4, width = 7.5)
```

# What fraction of the time is this effect observed?

```{r}
# Wrap just the necessary bits into a function
CreateCImodel <- function(state.size, effect.size){
  
  # Start with a new seed
  seed <- runif(1, min = 0, max = 1e7)
  set.seed(seed)
  
  # length of the series
  len <- 500
  treatment <- ceiling(2*len/3) #beginning of the treatment
  postperiod <- len - treatment
  
  # parameters of the state
  statedim <- 3
  T <- diag(statedim) #parameter matrix of state
  Q <- 0.01*diag(statedim) #variance of state
  x1 <- sin(2 * pi * seq_len(len) / 90)
  x2 <- sin(2 * pi * seq_len(len) / 360)
  
  # simulation of the state
  z0 <- c(5,0,0) #initial conditions: mu_0 = 5, beta_10 = 0, beta_20 = 0
  z <- matrix(ncol = statedim, nrow = len)
  
  # Start with a draw based on the initial conditions z0
  z[1,] <- z0%*%T + mvrnorm(1,rep(0,statedim), Q)
  
  # Update using the conditions from the last draw (previous column)
  for (i in 2:len) {
    z[i,] <- z[(i-1),] %*% T + mvrnorm(1, rep(0, statedim), Q)
  }
  
  # observations
  sigmasq <- 0.01 
  statesize <- state.size
  Z <- matrix(ncol = statedim, nrow = len) 
  Z[,1] <- rep(state.size, len)
  Z[,2] <- x1
  Z[,3] <- x2
  
  # Alter the state by the effect size
  e <- effect.size
  Ze <- Z
  Ze[treatment:len, 1] <- Ze[treatment:len, 1] + (statesize * e)
  
  # Create y vector with a state change (effect of intervention)
  ye <- vector(length = len)
  for (i in 1:len){
    ye[i] <- Ze[i,] %*% z[i,] + sqrt(sigmasq)*rnorm(1)
  }
  
  # Create data frame containing all variables to be plotted
  z.plot <- 
    Ze %>%
    data.frame %>%
    dplyr::rename(state = X1, x1 = X2, x2 = X3) %>%
    mutate("y" = ye) %>%
    mutate(index = seq(1, nrow(.))) %>%
    dplyr::select(y, x1, x2, index) %>%
    zoo
  
  return(z.plot)
}


pval <- c()
B <- 100
state.size <- 1
effect.size <- 0.25
for (i in 1:B) {
  impact <- CausalImpact(CreateCImodel(state.size, effect.size), 
                         pre.period = c(1, treatment), 
                         post.period = c(treatment + 1, 
                                         treatment + postperiod)
  )
  pval[i] <- impact$summary$p[2]
}

pwr <- sum(pval < 0.05) / length(pval)

pval %>%
  data.frame(pval = .) %>%
  ggplot(aes(pval)) +
  geom_density() +
  theme_bw() +
  annotate(geom = "text", 
           x = 0.3, 
           y = 3,
           label = paste0("Effect Size = ", effect.size, "\n",
                          "Power = ", pwr)) +
  ggsave(paste0("../figures/pvalue-density", effect.size, ".png"),
         height = 4, width = 7.5)
```



```{r}
n.iterations <- 100

state.size <- 1
effect.size <- 0.5
for (i in 1:B) {
  impact <- CausalImpact(CreateCImodel(state.size = state.size,
                                       effect.size = effect.size), 
                         pre.period = c(1, treatment), 
                         post.period = c(treatment + 1, 
                                         treatment + postperiod)
  )
  pval[i] <- impact$summary$p[2]
}

alpha <- 0.1
pwr <- sum(pval < alpha) / length(pval)

pval %>%
  data.frame(pval = .) %>%
  ggplot(aes(pval)) +
  geom_density() +
  theme_bw() +
  annotate(geom = "text", 
           x = max(pval) * 0.75, 
           y = 7,
           label = paste0("Effect Size = ", effect.size, "\n",
                          "Power = ", pwr)) +
  ggsave(paste0("../figures/pvalue-density", effect.size, ".png"),
         height = 4, width = 7.5)
```

